---
params:
  subtitle: ""
  title: ""
  authors: ""
title: '`r params$title`'
subtitle: '`r params$subtitle`'
author: '`r params$authors`'
date: 'Last run at: `r Sys.time()`'
output:
  bookdown::html_document2:
    self_contained: true
    fig_caption: yes
    code_folding: hide
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: TRUE
  bookdown::pdf_document2:
    fig_caption: yes
    number_sections: yes
  bookdown::word_document2:
    fig_caption: yes
    number_sections: yes
    toc: yes
    toc_depth: 2
    fig_width: 5
bibliography: '`r path.expand("~/github/dataknut/refs/refs.bib")`'
---

<hr>

>This fridayFagPacket was first published as a [blog](https://dataknut.wordpress.com/2020/10/16/retrofit-or-bust/)

<hr>

# fridayFagPackets

Numbers that could have been done on the back of one and should probably come with a similar health warning...

>Find out [more](https://dataknut.github.io/fridayFagPackets/).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(data.table)
library(ggplot2)
```

# It's the cats, stupid
Inspired by @giulio_mattioli's [recent paper on the car dependence of dog ownership](https://twitter.com/giulio_mattioli/status/1466361022747455492) we thought we'd take a look at [cats](https://twitter.com/giulio_mattioli/status/1466710752606179331) and residential energy demand. Why? Well people like to keep their cats warm but, more importantly, they also cut big holes in doors and/or windows to let the cats in and out. Hardly a thermally sealed envelope!

# What's the data?

For now we're using:

 * postcode sector level estimates of cat ownership in the UK. Does such a thing exist? [YEAH](https://t.co/ZEwaB5YEHI)!
 * LSOA level data on gas and electricity 'consumption' at LSOA/SOA level aggregated to postcode sectors

```{r loadData}
postcodes_dt <- data.table::fread("~/Dropbox/data/UK_postcodes/PCD_OA_LSOA_MSOA_LAD_AUG20_UK_LU.csv.gz")
postcodes_dt[, pcd_sector := tstrsplit(pcds, " ", keep = c(1))]
lsoa_DT <- postcodes_dt[, .(nPostcodes = .N), keyby = .(pcd_sector, lsoa11cd, ladnm, ladnmw)]

gas_dt <- data.table::fread("~/Dropbox/data/beis/subnationalGas/lsoaDom/LSOA_GAS_2019.csv.gz")
gas_dt[, lsoa11cd := `Lower Layer Super Output Area (LSOA) Code`]
gas_dt[, mean_gas_kWh := `Mean consumption (kWh per meter)`]
gas_dt[, total_gas_kWh := `Consumption (kWh)`]
gas_dt[, nGasMeters := `Number of consuming meters`]

elec_dt <- data.table::fread("~/Dropbox/data/beis/subnationalElec/lsoaDom/LSOA_ELEC_2019.csv.gz")
elec_dt[, lsoa11cd := `Lower Layer Super Output Area (LSOA) Code`]
elec_dt[, mean_elec_kWh := `Mean domestic electricity consumption 
(kWh per meter)`]
elec_dt[, total_elec_kWh := `Total domestic electricity consumption (kWh)`]
elec_dt[, nElecMeters := `Total number of domestic electricity meters`]

setkey(gas_dt, lsoa11cd)
setkey(elec_dt, lsoa11cd)
setkey(lsoa_DT, lsoa11cd)

merged_lsoa_DT <- gas_dt[, .(lsoa11cd, mean_gas_kWh, total_gas_kWh, nGasMeters)][elec_dt[, .(lsoa11cd,mean_elec_kWh,total_elec_kWh,nElecMeters)]][lsoa_DT]

# remove the record for postcodes which did not have a postcode sector
message("How many LSOAs do not map to a postcode sector?")
nrow(merged_lsoa_DT[is.na(pcd_sector)])
head(merged_lsoa_DT[is.na(pcd_sector)])

# !is.na(pcd_sector)
postcode_sector_energy <- merged_lsoa_DT[, .(nLSOAs = .N,
                                                               nPostcodes = sum(nPostcodes),
                                             mean_gas_kWh = mean(mean_gas_kWh, na.rm = TRUE),
                                             total_gas_kWh = sum(total_gas_kWh, na.rm = TRUE),
                                             mean_elec_kWh = mean(mean_elec_kWh, na.rm = TRUE),
                                             total_elec_kWh = sum(total_elec_kWh, na.rm = TRUE),
                                             nGasMeters = sum(nGasMeters, na.rm = TRUE),
                                             nElecMeters = sum(nElecMeters, na.rm = TRUE)), keyby = .(pcd_sector, ladnm, ladnmw)]
head(postcode_sector_energy)

# cats
cats_DT <- data.table::fread("~/Dropbox/data/UK_Animal and Plant Health Agency/APHA0372-Cat_Density_Postcode_District.csv")
cats_DT[, pcd_sector := PostcodeDistrict]

setkey(cats_DT, pcd_sector)
setkey(postcode_sector_energy, pcd_sector)

pc_district <- cats_DT[postcode_sector_energy]

```

We could also use @SERL_UK's [smart meter gas/elec data](https://twitter.com/dataknut/status/1466712963222540289?s=20), dwelling characteristics and pet ownership (but no species detail :-) 

# What do we find?

Well, in some places there seem to be a lot of estimated cats...

```{r maxCats}
pc_district[, mean_Cats := EstimatedCatPopulation/nElecMeters]
head(pc_district[, .(PostcodeDistrict, EstimatedCatPopulation, mean_Cats, nPostcodes, nElecMeters)][order(-mean_Cats)])
```
LL23 is on the south east corner of the [Snowdonia National Park...](https://www.google.co.uk/maps/place/Bala+LL23/@52.8953768,-3.775299,11z/data=!3m1!4b1!4m5!3m4!1s0x4865404ae1208f67:0x65a437b997c0dfb2!8m2!3d52.8825403!4d-3.6497989) while EH25 is on the outskirts of [Edinburgh](https://www.google.co.uk/maps/place/EH25/@55.8518992,-3.2076308,13z/data=!4m5!3m4!1s0x4887bf6548dd78d7:0xd6f980c5a3b93592!8m2!3d55.8560564!4d-3.1733124).

Is there a correlation between estimated total cats and the number of dwellings (electricity meters)?

```{r testTotalGas}
ggplot2::ggplot(pc_district, aes(x = nElecMeters , y = EstimatedCatPopulation)) +
  geom_point() +
  geom_smooth()
```

Is there a correlation between estimated cat ownership and energy use?

```{r testTotalGas}
ggplot2::ggplot(pc_district, aes(x = EstimatedCatPopulation, y = total_gas_kWh)) +
  geom_point()
```

```{r testMeanGas}

ggplot2::ggplot(pc_district, aes(x = mean_Cats, y = mean_gas_kWh)) +
  geom_point()
```
```{r testTotalElec}
ggplot2::ggplot(pc_district, aes(x = EstimatedCatPopulation, y = total_elec_kWh)) +
  geom_point()
```
```{r testMeanElec}
ggplot2::ggplot(pc_district, aes(x = mean_Cats, y = mean_elec_kWh)) +
  geom_point()
```

# R packages used

 * bookdown [@bookdown]
 * data.table [@data.table]
 * ggplot2 [@ggplot2]
 * knitr [@knitr]
 * rmarkdown [@rmarkdown]
 
# References
