---
params:
  subtitle: ""
  title: ""
  authors: ""
title: '`r params$title`'
subtitle: '`r params$subtitle`'
author: '`r params$authors`'
date: 'Last run at: `r Sys.time()`'
output:
  bookdown::html_document2:
    self_contained: true
    fig_caption: yes
    code_folding: hide
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: TRUE
  bookdown::pdf_document2:
    fig_caption: yes
    number_sections: yes
  bookdown::word_document2:
    fig_caption: yes
    number_sections: yes
    toc: yes
    toc_depth: 2
    fig_width: 5
bibliography: '`r path.expand("~/bibliography.bib")`'
---

<hr>

>This fridayFagPacket was first published at...

<hr>

# fridayFagPackets

Numbers that could have been done on the back of one and should probably come with a similar health warning...

>Find out [more](https://dataknut.github.io/fridayFagPackets/).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)

```

# It's the cats, stupid
Inspired by `@giulio_mattioli`'s [recent paper on the car dependence of dog ownership](https://twitter.com/giulio_mattioli/status/1466361022747455492) we thought we'd take a look at [cats](https://twitter.com/giulio_mattioli/status/1466710752606179331) and residential energy demand. Why? Well people like to keep their cats warm but, more importantly, they also cut big holes in doors and/or windows to let the cats in and out. Hardly a thermally sealed envelope!

# What's the data?

For now we're using:

 * postcode sector level estimates of cat ownership in the UK in 2015. Does such a thing exist? [YEAH](https://data.gov.uk/dataset/febd29ff-7e7d-4f82-9908-031f7f0e0860/cat-population-per-postcode-district)! "_This dataset gives the mean estimate for population for each district, and was generated as part of the delivery of commissioned research. The data contained within this dataset are modelled figures, based on national estimates for pet population, and available information on Veterinary activity across GB. The data are accurate as of 01/01/2015. The data provided are summarised to the postcode district level. Further information on this research is available in a research publication by James Aegerter, David Fouracre & Graham C. Smith, discussing the structure and density of pet cat and dog populations across Great Britain._"
 * LSOA level data on [gas](https://www.gov.uk/government/collections/sub-national-gas-consumption-data) and [electricity](https://www.gov.uk/government/collections/sub-national-electricity-consumption-data) 'consumption' at LSOA/SOA level for 2015 aggregated to postcode sectors
 * [Indices of Deprivation 2019](https://www.gov.uk/government/statistics/english-indices-of-deprivation-2019) for England

```{r loadCats}

# cats
cats_DT <- data.table::fread(paste0(dp, "UK_Animal and Plant Health Agency/APHA0372-Cat_Density_Postcode_District.csv"))
cats_DT[, pcd_sector := PostcodeDistrict]

setkey(cats_DT, pcd_sector)

nrow(cats_DT)

setkey(pc_sector_energy_dt, pcd_sector)

nrow(pc_sector_energy_dt)

pc_district <- merge(cats_DT, pc_sector_energy_dt , by = "pcd_sector") # keeps only postcode sectors where we have cat data

nrow(pc_district)
# there are postcode sectors with no electricity meters - for now we'll remove them
# pending further investigation

summary(pc_district)
table(pc_district$GOR10NM, pc_district$rgn)
```

We could also use `@SERL_UK`'s [smart meter gas/elec data](https://twitter.com/dataknut/status/1466712963222540289?s=20), dwelling characteristics and pet ownership (but no species detail :-) 

# What do we find?

Well, in some places there seem to be a lot of estimated cats...

(We calculated mean cats per household by dividing by the number of electricity meters - probably a reasonable proxy)

```{r maxCats}
pc_district[, mean_Cats := EstimatedCatPopulation/nElecMeters]
head(pc_district[, .(PostcodeDistrict, EstimatedCatPopulation, mean_Cats, nPostcodes, nElecMeters)][order(-mean_Cats)])
```
SA63 is in south west [Wales](https://www.google.co.uk/maps/place/Clarbeston+Road+SA63/@51.8852685,-4.9147384,12z/data=!3m1!4b1!4m5!3m4!1s0x4868d5805b12efe5:0xca42ee4bc84a2f77!8m2!3d51.8900045!4d-4.8502065) while LL23 is on the edge of the [Snowdonia National Park](https://www.google.co.uk/maps/place/Bala+LL23/@52.8953768,-3.7752989,11z/data=!3m1!4b1!4m5!3m4!1s0x4865404ae1208f67:0x65a437b997c0dfb2!8m2!3d52.8825403!4d-3.6497989)....

## More dwellings, more cats?

Is there a correlation between estimated total cats and the number of dwellings (electricity meters)?

```{r testTotalElecMeters}
ggplot2::ggplot(pc_district, aes(x = nElecMeters , y = EstimatedCatPopulation, 
                                 colour = GOR10NM)) +
  geom_point() +
  geom_smooth()
```

# More cats, more gas?

Is there a correlation between estimated cat ownership and total gas use?

```{r testTotalGas}
ggplot2::ggplot(pc_district, aes(x = EstimatedCatPopulation, y = total_gas_kWh,
                                 colour = GOR10NM)) +
  geom_point()
```

Or mean gas use and mean cats?

```{r testMeanGas}
pc_district[, mean_gas_kWh := total_gas_kWh/nGasMeters]
ggplot2::ggplot(pc_district, aes(x = mean_Cats, y = mean_gas_kWh, colour = GOR10NM)) +
  geom_point()
```

# More cats, more electricity?

Or total electricity use and cats?

```{r testTotalElec}
ggplot2::ggplot(pc_district, aes(x = EstimatedCatPopulation, y = total_elec_kWh, colour = GOR10NM)) +
  geom_point()
```

Or mean elec use and mean cats?

```{r testMeanElec}
pc_district[, mean_elec_kWh := total_elec_kWh/nGasMeters]
ggplot2::ggplot(pc_district, aes(x = mean_Cats, y = mean_elec_kWh, colour = GOR10NM)) +
  geom_point()
```

# More cats, more energy?

Or total energy use and total cats?

```{r testTotalEnergy}
pc_district[, total_energy_kWh := total_gas_kWh + total_elec_kWh]

ggplot2::ggplot(pc_district, aes(x = EstimatedCatPopulation, y = total_energy_kWh, colour = GOR10NM)) +
  geom_point() +
  geom_smooth()
```

Well, there may be something in there? Let's try a boxplot by cat deciles... Figure \@ref(fig:catDeciles)

```{r catDeciles, fig.cap = "Cat ownership deciles and total annual residenital electricity & gas use"}
pc_district[, cat_decile := dplyr::ntile(EstimatedCatPopulation, 10)]
#head(pc_district[is.na(cat_decile)])
ggplot2::ggplot(pc_district[!is.na(cat_decile)], aes(x = as.factor(cat_decile), y = total_energy_kWh/1000000)) +
  geom_boxplot() +
  facet_wrap(. ~ GOR10NM) +
  labs(x = "Cat ownership deciles",
       y = "Total domestic electricity & gas GWh",
       caption = "Postcode sectors (Data: BEIS & Animal and Plant Health Agency, 2015)")
```
Well...

```{r testMeanEnergy}
pc_district[, total_energy_kWh := total_gas_kWh + total_elec_kWh]
pc_district[, mean_energy_kWh := total_energy_kWh/nElecMeters]

ggplot2::ggplot(pc_district, aes(x = mean_Cats, y = mean_energy_kWh)) +
  geom_point()
```

# R packages used

 * bookdown [@bookdown]
 * data.table [@data.table]
 * ggplot2 [@ggplot2]
 * knitr [@knitr]
 * rmarkdown [@rmarkdown]
 
# References
